Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-04-20 co-regular-EPQ.tex
modify

\begin{restatable}{lemma}{priExecutionIsEnough}
\label{lemma:pri execution is enough}
Let $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$ and $e$ be a data-differentiated execution. Then, $e$ is not $\Gamma$-linearizable if and only if $e\vert_{\preceq p}$ is not $\Gamma$-linearizable, for some maximal priority $p$ in $e$.
\end{restatable}
\begin {proof} (Sketch)
To prove the $\textit{only if}$ direction, let $e$ be a data-differentiated execution linearizable w.r.t. $l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$. Since $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$ imposes no restriction on the operations in $u$, $v$ and $w$ with priorities incomparable to $p$, erasing all these operations results in a sequential execution which still satisfies this property. Similarly, for $\Gamma=\mathsf{UnmatchedMaxPriority}$.

@@

Thus, let $e$ be an execution such that $e\vert_{\preceq p}$ is linearizable w.r.t. $l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$. By the property above, we know that $e$ has a linearization $l' = u' \cdot \textit{put}(x,p) \cdot v' \cdot \textit{rm}(x) \cdot w'$, such that the projection of $l'$ to values of priority comparable to $p$ is $l$. 
%We can see that $u$, $v$ and $w$ are subsequences of $u'$, $v'$ an $w'$, respectively. 
Since $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$ does not have a condition on values of priority incomparable to $p$, we obtain that $l' \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$.
\end {proof}

@@

A differentiated execution $e$ is not $\mathsf{MatchedMaxPriority}$-linearizable when all the $\textit{put}$ operations in $e$ using the maximal priority $p$ are matched, and $e$ is not linearizable w.r.t. the set of sequential executions satisfying $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ for some value $x$ of priority $p$. We consider two cases depending on whether $e$ contains exactly one value with priority $p$ or at least two values. We denote by $\mathsf{MatchedMaxPriority}^{>}$ the strengthening of $\mathsf{MatchedMaxPriority}$ with the condition that all the values other than $x$ have a priority strictly smaller than $p$ (corresponding to the first case), and by $\mathsf{MatchedMaxPriority}^{=}$ the strengthening of the same formula with the negation of this condition (corresponding to the second case). 
We use particular instances of register automata~\cite{finite-memory, the paper I found} whose states include only two registers, one for storing a priority guessed at the initial state, and one for storing the priority of the current action in the execution. The transitions can check equality or the order relation $\prec$ between the values stored in the two registers. Instead of formalizing the full class of register automata, we consider a simpler class which suffices our needs. More precisely, we consider a class of labeled transition systems whose states consist of a finite control part and a register $r$ interpreted to elements of $\mathbb{P}$. The transition labels can be one of the following:
\begin{itemize}
	\item $r=*$ for storing an arbitrary value to $r$,
	\item $\textit{call}(\textit{rm},a)$ and $\textit{ret}(\textit{rm},a)$ for reading call/return actions of a remove,
	\item $\textit{call}(\textit{put},d,g)$ where $g\in\{=r,\prec r,true\}$ is a guard, for reading a call action $\textit{call}(\textit{put},d,p)$ of a put and checking whether $p$ is either equal to or smaller than the value stored in $r$, or arbitrary,
	\item $\textit{ret}(\textit{put},d,true)$ for reading a return action $\textit{ret}(\textit{put},d,p)$ for some $p$.
\end{itemize}
The set of words accepted by such a transition system can be defined as usual.

@@

\begin{figure}[htbp]
TODO UPDATE THE ARGUMENTS OF ACTIONS
  \centering
  \includegraphics[width=0.4 \textwidth]{figures/PIC-HIS-INTRO-GAP-EPQ1L.pdf}
%\vspace{-10pt}
  \caption{An execution that does not linearizable w.r.t $\mathsf{MatchedMaxPriority}^{>}$}
  \label{fig:introduce gap for EPQ1Lar}
\end{figure}

TODO GIVE THE FORMAL DEFINITION OF THE LEFT-RIGHT CONSTRAINT

In Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}, we introduce the notion of left-right constraint, which is a graph, and the existence of cycle though value with maximal priority in it formally modelled the case in \figurename~\ref{fig:introduce gap for EPQ1Lar}. For example, in \figurename~\ref{fig:introduce gap for EPQ1Lar}, $b$ is covered by $f$, $e$ and $d$. We then prove that getting rid of cycle though value with maximal priority in left-right constraint is enough for ensure linearizable w.r.t $\mathsf{MatchedMaxPriority}^{>}$.

When there is a cycle $d_1 \rightarrow \ldots \rightarrow d_m \rightarrow x \rightarrow d_1$ through value $x$ with maximal priority in left-right constraint, we say that $x$ is covered by $d_1,\ldots,d_m$. Assume that $x$ is covered by $d_1,\ldots,d_m$, we can safely rename $x$ into $b$, rename $d_1,\ldots,d_m$ into $a$ and rename all other value into $c$ by data-independence. Such execution can be recognized by witness automata, since between the first $\textit{ret}(\textit{put},a,\_)$ and the last $\textit{call}(\textit{rm},a)$ (if exists), $\textit{ret}(\textit{put},a,\_)$ and $\textit{call}(\textit{rm},a)$ occurs in pair and there is no need for count. In Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}, we construct a set $\textit{Auts}_{\textit{1-lar}}$ of register automata, and use $\textit{Auts}_{\textit{1-lar}}$ to prove that $\mathsf{MatchedMaxPriority}^{>}$ is co-regular, as stated by the following lemma.

\begin{restatable}{lemma}{EPQOneLarisCoRegular}
\label{lemma:EPQ1Lar is co-regular}
$\mathsf{MatchedMaxPriority}^{>}$ is co-regular.
\end{restatable}

TODO GIVE THE AUTOMATON

More details can be found in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}. TODO MOVE THIS TO THE RELATED WORK The notion of left-right constraint used for priority queue is inspired by left-right constraint of queue \cite{Bouajjani:2015}.

>>>>

\begin{restatable}{lemma}{priExecutionIsEnough}
\label{lemma:pri execution is enough}
Let $\Gamma\in \{\mathsf{UnmatchedMaxPriority}, \mathsf{MatchedMaxPriority}\}$ and $e$ be a data-differentiated execution. Then, $e$ is not $\Gamma$-linearizable if and only if $e\vert_{\preceq p}$ is not $\Gamma$-linearizable, for {\color{blue}any} maximal priority $p$ in $e$.
\end{restatable}
\begin {proof} (Sketch) 
{\color {blue}We prove an equivalent property: $e$ is $\Gamma$-linearizable if and only if $e\vert_{\preceq p}$ is $\Gamma$-linearizable for some maximal priority $p$ in $e$.} %We consider the case of $\mathsf{MatchedMaxPriority}$ and the case of $\mathsf{UnmatchedMaxPriority}$ is similar.

To prove the $\textit{only if}$ direction, let $e$ be a data-differentiated execution linearizable w.r.t. $l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,x)$. Since $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,x)$ imposes no restriction on the operations in $u$, $v$ and $w$ with priorities incomparable to $p$, erasing all these operations results in a sequential execution which still satisfies this property. Similarly, for $\Gamma=\mathsf{UnmatchedMaxPriority}$.

@@

Thus, let $e$ be an execution such that $e\vert_{\preceq p}$ is linearizable w.r.t. $l = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,x)$. By the property above, we know that $e$ has a linearization $l' = u' \cdot \textit{put}(x,p) \cdot v' \cdot \textit{rm}(x) \cdot w'$, such that the projection of $l'$ to values of priority comparable to $p$ is $l$.
%We can see that $u$, $v$ and $w$ are subsequences of $u'$, $v'$ an $w'$, respectively.
Since $\mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,x)$ does not have a condition on values of priority incomparable to $p$, we obtain that $l' \in \mathsf{MatchedMaxPriority}\mathsf{\text{-}Seq}(s,\alpha)$.
\end {proof}

@@

A differentiated execution $e$ is not $\mathsf{MatchedMaxPriority}$-linearizable when all the $\textit{put}$ operations in $e$ using the maximal priority $p$ are matched, and $e$ is not linearizable w.r.t. the set of sequential executions satisfying $\mathsf{MatchedMaxPriority\text{-}Seq}(e,x)$ for {\color {blue}any value $x$ of priority $p$}. We consider two cases depending on whether $e$ contains exactly one value with priority $p$ or at least two values. We denote by $\mathsf{MatchedMaxPriority}^{>}$ the strengthening of $\mathsf{MatchedMaxPriority}$ with the condition that all the values other than $x$ have a priority strictly smaller than $p$ (corresponding to the first case), and by $\mathsf{MatchedMaxPriority}^{=}$ the strengthening of the same formula with the negation of this condition (corresponding to the second case).
We use particular instances of register automata~\cite{finite-memory, the paper I found} whose states include only two registers, one for storing a priority guessed at the initial state, and one for storing the priority of the current action in the execution. The transitions can check equality or the order relation $\prec$ between the values stored in the two registers. Instead of formalizing the full class of register automata, we consider a simpler class which suffices our needs. More precisely, we consider a class of labeled transition systems whose states consist of a finite control part and a register $r$ interpreted to elements of $\mathbb{P}$. The transition labels can be one of the following:
\begin{itemize}
	\item $r=*$ for storing an arbitrary value to $r$,
	\item $\textit{call}(\textit{rm},a)$ and $\textit{ret}(\textit{rm},a)$ for reading call/return actions of a remove,
	\item $\textit{call}(\textit{put},d,g)$ where $g\in\{=r,\prec r,true\}$ is a guard, for reading a call action $\textit{call}(\textit{put},d,p)$ of a put and checking whether $p$ is either equal to or smaller than the value stored in $r$, or arbitrary,
	\item $\textit{ret}(\textit{put},d,true)$ for reading a return action $\textit{ret}(\textit{put},d,p)$ for some $p$.
\end{itemize}
The set of words accepted by such a transition system can be defined as usual.

@@

\begin{figure}[htbp]
TODO UPDATE THE ARGUMENTS OF ACTIONS
  \centering
  \includegraphics[width=0.4 \textwidth]{figures/PIC-HIS-INTRO-GAP-EPQ1L.pdf}
%\vspace{-10pt}
  \caption{An execution that is not $\mathsf{MatchedMaxPriority}^{>}$-linearizable}
  \label{fig:introduce gap for EPQ1Lar}
\end{figure}

TODO GIVE THE FORMAL DEFINITION OF THE LEFT-RIGHT CONSTRAINT
{\color {blue}Let us introduce left-right constraint to detect the phenomenon in \figurename~\ref{fig:introduce gap for EPQ1Lar}.

\begin{definition}\label{def:left-right constraint for matched put and rm operations}
Given a data-differentiated $e$ without $\textit{rm}(\textit{empty})$, and let $\textit{put}(x,p)$ and $\textit{rm}(x)$ be the only operations with priority $p$. The left-right constraint of $\textit{put}(x,p)$ and $\textit{rm}(x)$ is the graph $G$ where:

\begin{itemize}
\item the nodes are the values of $e$, to which we add a node,
\item there is an edge from $d_1$ to $x$, if $\textit{put}(d_1,\_) <_{\textit{hb}} \textit{put}(x,p)$ or $\textit{rm}(x)$,
\item there is an edge from $x$ to $d_1$, if $\textit{rm}(x)$ happens before $\textit{rm}(d_1)$ or $\textit{rm}(d_1)$ does not exists,
\item there is an edge from $d_1$ to $d_2$, if $\textit{put}(d_1,\_) <_{\textit{hb}} \textit{rm}(d_2,\_)$.
\end{itemize} 
\end{definition}

The existence of cycle though value with maximal priority in left-right constraint formally modelled the phenomenon in \figurename~\ref{fig:introduce gap for EPQ1Lar}. For example, in \figurename~\ref{fig:introduce gap for EPQ1Lar}, $f \rightarrow e \rightarrow d \rightarrow b \rightarrow f$ is a cycle in left-right constraint. We then prove that getting rid of cycle though value with maximal priority in left-right constraint is enough for ensure linearizable w.r.t $\mathsf{MatchedMaxPriority}^{>}$.
}

When there is a cycle $d_1 \rightarrow \ldots \rightarrow d_m \rightarrow x \rightarrow d_1$ through value $x$ with maximal priority in left-right constraint, we say that $x$ is covered by $d_1,\ldots,d_m$. Assume that $x$ is covered by $d_1,\ldots,d_m$, we can safely rename $x$ into $b$, rename $d_1,\ldots,d_m$ into $a$ and rename all other value into $c$ by data-independence. Such execution can be recognized by register automata, since between the first $\textit{ret}(\textit{put},a,\_)$ and the last $\textit{call}(\textit{rm},a)$ (if exists), $\textit{ret}(\textit{put},a,\_)$ and $\textit{call}(\textit{rm},a)$ occurs in pair and there is no need for count. In Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}, we construct a set $\textit{Auts}_{\textit{1-lar}}$ of register automata, {\color {blue}and prove that $\textit{Auts}_{\textit{1-lar}}$ is $\mathsf{MatchedMaxPriority}^{>}$-complete, as stated by the following lemma.

\begin{restatable}{lemma}{EPQOneLarisCoRegular}
\label{lemma:EPQ1Lar is co-regular}
$\textit{Auts}_{\textit{1-lar}}$ is $\mathsf{MatchedMaxPriority}^{>}$-complete.
\end{restatable}
}

{\color {blue}
For example, to capture non-$\mathsf{MatchedMaxPriority}^{>}$-linearizable execution $e$ where $e \vert_{b} = \textit{call}(\textit{put},b,\_) \cdot \textit{ret}(\textit{put},b,\_) \cdot \textit{call}(\textit{rm},b) \cdot \textit{ret}(\textit{rm},b)$, we use the register automaton in \figurename~\ref{fig:automata APQ1Lar-1 in paper}. Here $c_1 = c + \textit{ret}(\textit{rm},a)$, $c_2 = c + \textit{call}(\textit{put},a,=r)$, $c_3 = c_2 + \textit{ret}(\textit{rm},a)$, where $c = \textit{call}(\textit{put},d,\top),\textit{ret}(\textit{put},d,\top), \textit{call}(\textit{rm},d)$, $\textit{ret}(\textit{rm},d),\textit{call}(\textit{rm},\textit{empty}),\textit{ret}(\textit{rm},\textit{empty})$. There are three paths in $\mathcal{A}_{\textit{l-lar}}^1$, since there are three possible position of  the first $\textit{ret}(\textit{put},a)$ in $e$.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1 \textwidth]{figures/PIC_AUTO_PQ1Lar-pprr.pdf}
%\vspace{-10pt}
  \caption{Automaton $\mathcal{A}_{\textit{l-lar}}^1$}
  \label{fig:automata APQ1Lar-1 in paper}
\end{figure}
}
TODO GIVE THE AUTOMATON

More details can be found in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}. %TODO MOVE THIS TO THE RELATED WORK

### 2017-04-20 introduction.tex
modify

\noindent {\bf Related work}
\noindent The most related work of our paper is ~\cite{DBLP:conf/icalp/BouajjaniEEH15}. Compared to ~\cite{DBLP:conf/icalp/BouajjaniEEH15}, our work are specific to priority queue and thus have points not occur in queues and stacks. Due to the partial-order priorities, to prove the correctness of recursive procedure, we need to alter the positions of linearization points of incomparable items while keeps the linearization points of comparable items unchanged. For checking violations of one local property, we just need to find a value, such that at every possible time point to locate its remove operation, there is already values with smaller priorities in priority queue. For checking violations of another local property, we just need to find two values with maximal priority, such that every possible time point to locate remove operation of one value is disabled by remove operation of another value. Such phenomenons are unique for priority queue.

>>>>

\noindent {\bf Related work}
\noindent The most related work of our paper is ~\cite{DBLP:conf/icalp/BouajjaniEEH15}. Compared to ~\cite{DBLP:conf/icalp/BouajjaniEEH15}, our work are specific to priority queue and thus have points not occur in queues and stacks. Due to the partial-order priorities, to prove the correctness of recursive procedure, we need to alter the positions of linearization points of incomparable items while keeps the linearization points of comparable items unchanged. {\color {blue}In the proof process we use the notion of left-right constraint, which is inspired by left-right constraint of queue \cite{Bouajjani:2015}.} For checking violations of one local property, we just need to find a value, such that at every possible time point to locate its remove operation, there is already values with smaller priorities in priority queue. For checking violations of another local property, we just need to find two values with maximal priority, such that every possible time point to locate remove operation of one value is disabled by remove operation of another value. Such phenomenons are unique for priority queue.
